Unbalanced Mallows Models for Optimizing Expensive Black-Box Permutation Problems

### 2021-04-16 paper/main.tex
paper
fix double label
update
fix margin

\myparagraph{Datasets}
%
%For the LOP, we generated synthetic instances as follows. First, we create a sample $S$ of 200 permutations of size $n=20$ from a MM with $\sigma_0$ equal to the identity permutation and $\theta=-\log(\textit{Uniformity})$, where  $\textit{Uniformity}\in\{0.5,$ $0.7,$ $0.9\}$, i.e., from peaked distributions to almost uniform distributions. Then, we generate matrix $a$ as $a_{i,j} = \sum_{\sigma\in S}\sum_{i<j}\mathbb{I}\{\sigma(i)<\sigma(j)\}$, where ${\mathbb{I}\colon S \times S \to \{0,1\}}$ is the indicator function. In principle, these should be best-case instances for UMM since their fitness landscape closely matches the Mallows model.%\MANUEL{Ekhine: Could you revise this text? I wanted to explain why we use synthetic datasets.}
We consider real LOP datasets from the LOLIB
repository.\footnote{\url{http://grafo.etsii.urjc.es/optsicom/lolib/}} In
particular, we have taken two instances of various types focusing on relatively
small instances, since he computation time of CEGO grows rapidly with instance
size: \texttt{N-p40-01}, \texttt{N-p40-02}, \texttt{N-p50-01} and
\texttt{N-p50-02} of type \emph{RandB}; \texttt{N-t59d11xx} and
\texttt{N-t59b11xx} of type \emph{IO}; and \texttt{N-sgb75.01} and
\texttt{N-sgb75.02} of type \emph{SGB}.
% Random instances of type B of various sizes $n \in \{50, 100, 150, 200,
% 250\}$. These are harder problems than the synthetic instances, not only
% because of the larger search space, but also because of the different instance
% types.
In the case of the PFSP, we consider the same instances as
\citet{ZaeStoBar2014:ppsn}, i.e., \texttt{reC05}, \texttt{reC13},
\texttt{reC19} and \texttt{rec31}, with $n \in \{20, 20, 30, 50\}$ and $M \in \{5, 20, 10, 15\}$, respectively.

@@

\afterpage{
\begin{figure*}
\small
\centering%
\begin{minipage}{0.49\linewidth}
  \texttt{N-p40-01} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{./img/fitness_lop_RandB_N-p40-01}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{N-p40-02} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{./img/fitness_lop_RandB_N-p40-02}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{N-t59b11xx} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{./img/fitness_lop_IO_N-t59b11xx}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{N-t59d11xx} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{./img/fitness_lop_IO_N-t59d11xx}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{SGB\_N-sgb75\_01} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{./img/fitness_lop_SGB_N-sgb75_01}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{SGB\_N-sgb75\_01} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{./img/fitness_lop_SGB_N-sgb75_02}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{reC05} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{./img/fitness_pfsp_rec05_txt}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{reC13} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{./img/fitness_pfsp_rec13_txt}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{reC19} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{./img/fitness_pfsp_rec19_txt}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{reC31} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{./img/fitness_pfsp_rec31_txt}
\end{minipage}
\label{fig:results}
\caption{Mean fitness  (and standard deviation)  of each solution evaluated by each algorithm on various instances.\label{fig:results}}
\vspace{10em}
\end{figure*}
}

@@

\newcommand{\mcolc}[2]{\multicolumn{#1}{c}{\bf #2}}
\afterpage{%
\begin{table*}
  \caption{Mean fitness (and standard deviation) of the best solution found at the end of each run, over 10 independent runs for each problem instance. The 95\% confidence interval corresponds to the two independent samples t-test on the mean difference between the fitness of \CEGOinv minus the one of UMM$_{400}$. 
  Mean runtime is shown in hours. 
  %CEGO runs were stopped after 144 hours. This was the case for all LOP instances.
  \label{tab:results}}
 % \resizebox{\textwidth}{!}{%
\begin{tabular}{r@{\hskip -1ex}r*{2}{r@{\hskip 1ex}l}r@{\hskip 0.5ex}lr@{\hskip 0.75ex}r}
 \toprule
            &              & \mcolc{4}{Mean fitness (sd)} & \mcolc{2}{\multirow[b]{2.45}{*}{\shortstack{95\% CI of the                                                                                       \\mean difference}}} & \mcolc{2}{Mean runtime} \\\cmidrule(lr){3-6}\cmidrule(lr){9-10}
\bf Problem & \bf Instance & \mcolc{2}{\CEGOinv}              & \mcolc{2}{UMM$_{400}$}                                                                  &           &           & \bf \CEGOinv                  & \bf UMM$_{400}$    \\\midrule
LOP         & N-p40-01     & 10496.6                      & (116.8)                                                                         & 12369.1   & (389.9)   & {[}-2157.1,& -1587.9{]}    & 19.9 & 0.3 \\
         & N-p40-02     & 11714.7                      & (84.7)                                                                          & 13160.5   & (454.8)   & {[}-1773.5,& -1118.1{]}    & 20.1 & 0.3 \\
         & N-t59b11xx   & 102408.0                     & (3800.2)                                                                        & 114075.5  & (5389.2)  & {[}-16084.2,& -7250.8{]}   & 21.1 & 0.4 \\
         & N-t59d11xx   & 12692.2                      & (1097.9)                                                                        & 16662.9   & (2155.3)  & {[}-5618.5,& -2322.9{]}    & 21.2 & 0.4 \\
         & N-sgb75.01   & 1893533.2                    & (28594.6)                                                                       & 1992059.9 & (40235.7) & {[}-131577.1,& -65476.3{]} & 33.2 & 1.0 \\
         & N-sgb75.02   & 1895801.8                    & (82847.9)                                                                       & 1967132.3 & (49228.1) & {[}-136421.2,& -6239.8{]}  & 33.3 & 1.1 \\
         \midrule
PFSP        & rec05        & 1276.2                       & (15.1)                                                                          & 1276.3    & (8.4)     & {[}-11.8,& 11.6{]}         & 18.3 & 0.1 \\
       & rec13        & 1992.9                       & (27.6)                                                                          & 2057.8    & (22.8)    & {[}-88.8,& -41.0{]}        & 18.5 & 0.1 \\
       & rec19        & 2264.5                       & (45.0)                                                                          & 2304.8    & (37.7)    & {[}-79.4,& -1.2{]}         & 18.2 & 0.2 \\
       & rec31        & 3415.5                       & (62.1)                                                                          & 3419.3    & (37.0)    & {[}-52.6,& 45.0{]}         & 23.1 & 0.5 \\
\bottomrule
\end{tabular}
%}
\end{table*}

\begin{table*}
    \caption{Mean fitness, over 10 independent runs for each problem instance, of the best solution found at the end of each run of the original CEGO (\CEGOorig) and its improved version (\CEGOinv) and three variants of UMM that differ in the budget of evaluations $m \in \{100, 200, 400\}$. The runs of CEGO used a budget of $m=400$.\label{tab:results_fitness}}
\begin{tabular}{llrrrrr}
\toprule
\bf Problem & \bf   Instance &  \CEGOinv &  \CEGOorig &  UMM$_{400}$ &   UMM$_{200}$ &  UMM$_{100}$ \\
\midrule
    LOP &    N-p40-01 &    10496.6 &    15027.7 &    12453.7 &  1.336170e+04 &    14649.9 \\
    &    N-p40-02 &    11714.7 &    15150.2 &    13191.9 &  1.420900e+04 &    15206.2 \\
    &  N-t59b11xx &   102408.0 &   126204.0 &   114726.5 &  1.202020e+05 &   132039.5 \\
    &  N-t59d11xx &    12692.2 &    30199.8 &    16775.4 &  2.352089e+04 &    29884.1 \\
    &  N-sgb75.01 &  1893533.2 &  2095899.6 &  1983521.2 &  2.004006e+06 &  2046895.1 \\
    &  N-sgb75.02 &  1895801.8 &  2069286.9 &  1973083.9 &  2.001244e+06 &  2039356.1 \\
    \midrule
    PFSP &       rec05 &     1276.2 &     1318.7 &     1282.1 &  1.287000e+03 &     1320.2 \\
    &       rec13 &     1992.9 &     2096.0 &     2077.7 &  2.108500e+03 &     2131.6 \\
    &       rec19 &     2264.5 &     2328.4 &     2318.5 &  2.352100e+03 &     2386.5 \\
    &       rec31 &     3415.5 &     3512.3 &     3420.5 &  3.490900e+03 &     3474.3 \\
\bottomrule
\end{tabular}
\end{table*}

\begin{table*}
  \caption{Mean runtime  in minutes, over 10 independent runs for each problem instance, of the original CEGO (\CEGOorig) and its improved version (\CEGOinv) and three variants of UMM that differ in the budget of evaluations $m \in \{100, 200, 400\}$. The runs of CEGO used a budget of $m=400$.\label{tab:results_time}}
\begin{tabular}{llrrrrr}
\toprule
\bf Problem & \bf   Instance & \CEGOinv & \CEGOorig & UMM$_{400}$ & UMM$_{200}$ & UMM$_{100}$ \\
\midrule
    LOP     & N-p40-01       & 1195.90  & 1170.81   & 3.12        & 0.85    & 0.21        \\
            & N-p40-02       & 1207.24  & 1183.09   & 3.11        & 0.83    & 0.20        \\
            & N-t59b11xx     & 1265.98  & 1240.36   & 3.78        & 0.95    & 0.23        \\
            & N-t59d11xx     & 1269.93  & 1252.89   & 3.93        & 0.97    & 0.25        \\
            & N-sgb75.01     & 1994.22  & 2004.70   & 10.88       & 2.73    & 0.72        \\
            & N-sgb75.02     & 1995.09  & 1975.41   & 10.88       & 2.76    & 0.69        \\
\midrule
    PFSP    & rec05          & 1099.63  & 917.81    & 0.80        & 0.20    & 0.10        \\
            & rec13          & 1109.51  & 1043.08   & 0.85        & 0.20    & 0.10        \\
            & rec19          & 1090.06  & 1087.38   & 1.76        & 0.47    & 0.10        \\
            & rec31          & 1387.09  & 1367.26   & 4.77        & 1.28    & 0.32        \\
\bottomrule
\end{tabular}
\end{table*}

>>>>

\myparagraph{Datasets}
%
%For the LOP, we generated synthetic instances as follows. First, we create a sample $S$ of 200 permutations of size $n=20$ from a MM with $\sigma_0$ equal to the identity permutation and $\theta=-\log(\textit{Uniformity})$, where  $\textit{Uniformity}\in\{0.5,$ $0.7,$ $0.9\}$, i.e., from peaked distributions to almost uniform distributions. Then, we generate matrix $a$ as $a_{i,j} = \sum_{\sigma\in S}\sum_{i<j}\mathbb{I}\{\sigma(i)<\sigma(j)\}$, where ${\mathbb{I}\colon S \times S \to \{0,1\}}$ is the indicator function. In principle, these should be best-case instances for UMM since their fitness landscape closely matches the Mallows model.%\MANUEL{Ekhine: Could you revise this text? I wanted to explain why we use synthetic datasets.}
We consider LOP datasets from the LOLIB
repository.\footnote{\url{http://grafo.etsii.urjc.es/optsicom/lolib/}} In
particular, we have taken two instances of various types focusing on relatively
small instances, since he computation time of CEGO grows rapidly with instance
size: \texttt{N-p40-01}, \texttt{N-p40-02}, \texttt{N-p50-01} and
\texttt{N-p50-02} of type \emph{RandB}; \texttt{N-t59d11xx} and
\texttt{N-t59b11xx} of type \emph{IO}; and \texttt{N-sgb75.01} and
\texttt{N-sgb75.02} of type \emph{SGB}.
% Random instances of type B of various sizes $n \in \{50, 100, 150, 200,
% 250\}$. These are harder problems than the synthetic instances, not only
% because of the larger search space, but also because of the different instance
% types.
In the case of the PFSP, we consider the same instances as
\citet{ZaeStoBar2014:ppsn}, i.e., \texttt{reC05}, \texttt{reC13},
\texttt{reC19} and \texttt{rec31}, with $n \in \{20, 20, 30, 50\}$ and $M \in \{5, 20, 10, 15\}$, respectively.

@@

\afterpage{
\begin{figure*}
\small
\centering%
\begin{minipage}{0.49\linewidth}
  \texttt{N-p40-01} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{./img/fitness_lop_RandB_N-p40-01}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{N-p40-02} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{./img/fitness_lop_RandB_N-p40-02}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{N-t59b11xx} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{./img/fitness_lop_IO_N-t59b11xx}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{N-t59d11xx} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{./img/fitness_lop_IO_N-t59d11xx}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{SGB\_N-sgb75\_01} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{./img/fitness_lop_SGB_N-sgb75_01}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{SGB\_N-sgb75\_01} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{./img/fitness_lop_SGB_N-sgb75_02}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{reC05} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{./img/fitness_pfsp_rec05_txt}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{reC13} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{./img/fitness_pfsp_rec13_txt}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{reC19} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{./img/fitness_pfsp_rec19_txt}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{reC31} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{./img/fitness_pfsp_rec31_txt}
\end{minipage}
\caption{Mean fitness  (and standard deviation)  of each solution evaluated by each algorithm on various instances.\label{fig:results}}
\vspace{10em}
\end{figure*}
}

@@

\newcommand{\mcolc}[2]{\multicolumn{#1}{c}{\bf #2}}
\afterpage{%
\begin{table*}
%\smaller
  \caption{Mean fitness (and standard deviation) of the best solution found at the end of each run, over 10 independent runs for each problem instance. The 95\% confidence interval corresponds to the two independent samples t-test on the mean difference between the fitness of \CEGOinv minus the one of UMM$_{400}$. 
  Mean runtime is shown in hours. 
  %CEGO runs were stopped after 144 hours. This was the case for all LOP instances.
  \label{tab:results}}
 % \resizebox{\textwidth}{!}{%
\begin{tabular}{r@{\hskip -1ex}r*{2}{r@{\hskip 1ex}l}r@{\hskip 0.5ex}lr@{\hskip 0.75ex}r}
 \toprule
            &              & \mcolc{4}{Mean fitness (sd)} & \mcolc{2}{\multirow[b]{2.45}{*}{\shortstack{95\% CI of the                                                                                       \\mean difference}}} & \mcolc{2}{Mean runtime} \\\cmidrule(lr){3-6}\cmidrule(lr){9-10}
\bf Problem & \bf Instance & \mcolc{2}{\CEGOinv}              & \mcolc{2}{UMM$_{400}$}                                                                  &           &           & \bf \CEGOinv                  & \bf UMM$_{400}$    \\\midrule
LOP         & N-p40-01     & 10496.6                      & (116.8)                                                                         & 12369.1   & (389.9)   & {[}-2157.1,& -1587.9{]}    & 19.9 & 0.3 \\
         & N-p40-02     & 11714.7                      & (84.7)                                                                          & 13160.5   & (454.8)   & {[}-1773.5,& -1118.1{]}    & 20.1 & 0.3 \\
         & N-t59b11xx   & 102408.0                     & (3800.2)                                                                        & 114075.5  & (5389.2)  & {[}-16084.2,& -7250.8{]}   & 21.1 & 0.4 \\
         & N-t59d11xx   & 12692.2                      & (1097.9)                                                                        & 16662.9   & (2155.3)  & {[}-5618.5,& -2322.9{]}    & 21.2 & 0.4 \\
         & N-sgb75.01   & 1893533.2                    & (28594.6)                                                                       & 1992059.9 & (40235.7) & {[}-131577.1,& -65476.3{]} & 33.2 & 1.0 \\
         & N-sgb75.02   & 1895801.8                    & (82847.9)                                                                       & 1967132.3 & (49228.1) & {[}-136421.2,& -6239.8{]}  & 33.3 & 1.1 \\
         \midrule
PFSP        & rec05        & 1276.2                       & (15.1)                                                                          & 1276.3    & (8.4)     & {[}-11.8,& 11.6{]}         & 18.3 & 0.1 \\
       & rec13        & 1992.9                       & (27.6)                                                                          & 2057.8    & (22.8)    & {[}-88.8,& -41.0{]}        & 18.5 & 0.1 \\
       & rec19        & 2264.5                       & (45.0)                                                                          & 2304.8    & (37.7)    & {[}-79.4,& -1.2{]}         & 18.2 & 0.2 \\
       & rec31        & 3415.5                       & (62.1)                                                                          & 3419.3    & (37.0)    & {[}-52.6,& 45.0{]}         & 23.1 & 0.5 \\
\bottomrule
\end{tabular}
%}
\end{table*}

\begin{table*}
%\smaller
    \caption{Mean fitness, over 10 independent runs for each problem instance, of the best solution found at the end of each run of the original CEGO (\CEGOorig) and its improved version (\CEGOinv) and three variants of UMM that differ in the budget of evaluations $m \in \{100, 200, 400\}$. The runs of CEGO used a budget of $m=400$.\label{tab:results_fitness}}
\begin{tabular}{llrrrrr}
\toprule
\bf Problem & \bf   Instance &  \CEGOinv &  \CEGOorig &  UMM$_{400}$ &   UMM$_{200}$ &  UMM$_{100}$ \\
\midrule
    LOP &    N-p40-01 &    10496.6 &    15027.7 &    12453.7 &  1.336170e+04 &    14649.9 \\
    &    N-p40-02 &    11714.7 &    15150.2 &    13191.9 &  1.420900e+04 &    15206.2 \\
    &  N-t59b11xx &   102408.0 &   126204.0 &   114726.5 &  1.202020e+05 &   132039.5 \\
    &  N-t59d11xx &    12692.2 &    30199.8 &    16775.4 &  2.352089e+04 &    29884.1 \\
    &  N-sgb75.01 &  1893533.2 &  2095899.6 &  1983521.2 &  2.004006e+06 &  2046895.1 \\
    &  N-sgb75.02 &  1895801.8 &  2069286.9 &  1973083.9 &  2.001244e+06 &  2039356.1 \\
    \midrule
    PFSP &       rec05 &     1276.2 &     1318.7 &     1282.1 &  1.287000e+03 &     1320.2 \\
    &       rec13 &     1992.9 &     2096.0 &     2077.7 &  2.108500e+03 &     2131.6 \\
    &       rec19 &     2264.5 &     2328.4 &     2318.5 &  2.352100e+03 &     2386.5 \\
    &       rec31 &     3415.5 &     3512.3 &     3420.5 &  3.490900e+03 &     3474.3 \\
\bottomrule
\end{tabular}
\end{table*}

\begin{table*}
%\smaller
  \caption{Mean runtime  in minutes, over 10 independent runs for each problem instance, of the original CEGO (\CEGOorig) and its improved version (\CEGOinv) and three variants of UMM that differ in the budget of evaluations $m \in \{100, 200, 400\}$. The runs of CEGO used a budget of $m=400$.\label{tab:results_time}}
\begin{tabular}{llrrrrr}
\toprule
\bf Problem & \bf   Instance & \CEGOinv & \CEGOorig & UMM$_{400}$ & UMM$_{200}$ & UMM$_{100}$ \\
\midrule
    LOP     & N-p40-01       & 1195.90  & 1170.81   & 3.12        & 0.85    & 0.21        \\
            & N-p40-02       & 1207.24  & 1183.09   & 3.11        & 0.83    & 0.20        \\
            & N-t59b11xx     & 1265.98  & 1240.36   & 3.78        & 0.95    & 0.23        \\
            & N-t59d11xx     & 1269.93  & 1252.89   & 3.93        & 0.97    & 0.25        \\
            & N-sgb75.01     & 1994.22  & 2004.70   & 10.88       & 2.73    & 0.72        \\
            & N-sgb75.02     & 1995.09  & 1975.41   & 10.88       & 2.76    & 0.69        \\
\midrule
    PFSP    & rec05          & 1099.63  & 917.81    & 0.80        & 0.20    & 0.10        \\
            & rec13          & 1109.51  & 1043.08   & 0.85        & 0.20    & 0.10        \\
            & rec19          & 1090.06  & 1087.38   & 1.76        & 0.47    & 0.10        \\
            & rec31          & 1387.09  & 1367.26   & 4.77        & 1.28    & 0.32        \\
\bottomrule
\end{tabular}
\end{table*}
